Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-12 latex/main.tex
Write up the parsimony analysis

>>>>

\subsection{Parsimony comparison of sc2ts and UShER}
\label{sec:supplement_sc2ts_usher_parsimony}
We performed an indepth analysis of the sc2ts and UShER ARGs 
(we refer to the UShER tree as an ARG for simplicity here as a single 
tree is also an ARG) in terms of parsimony.
A direct comparison is difficult because of the different aligments 
used as input, the different sets of sites used for inference and 
the fact that we are including deletions for a subset of sites in
the sc2ts ARG.

When we compare the intersection of the base sc2ts ARG (pre 
site-remapping), sc2ts is very slightly more parsimonious 
% TODO: https://github.com/jeromekelleher/sc2ts-paper/issues/684
with XX mutations over the YYY sites shared between them.
After site remapping, and including all mutations involving 
deletion characters, sc2ts has 143,690 mutations more 
(2,078,370 vs 1,934,680). If we exclude mutations involving 
deletions in sc2ts, this is reduced to 46,371 more (2.3\%)
than UShER.
If we restrict to the 26,192 sites that were not remapped, 
sc2ts has 4,785 (0.26\%) fewer mutations than UShER,
and has around 2.5 times more mutations than UShER on the 
163 sites that were post-hoc remapped.

To provide a more objective measure of the overall parsimony 
of the inferred ARGs with respect to the Viridian alignments,
we remapped all 29903 sites in both ARGs 
using Fitch-Hartigan parsimony. Here the nucleotides
ACGT and ``-'' were treated as distinct characters, with all 
other values regarded as missing data. In this case, the sc2ts 
ARG has 629,864 (5.5\%) fewer mutations (10,840,687 
vs 11,470,551). Some of this difference may be due to the 
post-hoc incorporation of phylogenetic signal from deletions 
by sc2ts, but more work is required to fully understand 
the differences between the two approaches.

See notebook
\href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/analysis\_usher\_sc2ts.ipynb}{analysis\_usher\_sc2ts.ipynb}
for details.
